Modal Abstractions for Virtualizing Memory Addresses

### 2025-07-24 examples_C.tex
finish main pass through experiments

\begin{figure}\footnotesize
  \begin{lstlisting}[language=C,mathescape,escapeinside={(*}{*)}]
/* @param entry: virtual address of a non-present entry to initialize with a new physical page */
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \lnot(\textsf{entry\_present entryv}) \ast \textsf{entryp+KERNBASE} \mapsto_{\textsf{vpte,1}} \textsf{entryp entryv} \ast \ghostmaptoken{\textsf{id}}{(\mathsf{entryp})}{\textsf{level}} }$
void pte_initialize(pte_t *entry) {
  pte_t *local = kalloc(); // Allocate a full zeroed page for 512 8-byte entries(*\label{line:call_to_kallocC}*)
  entry->pfn = PTE_ADDR_TO_PFN((uintptr_t) local);(*\label{line:kalloc_install}*)
  $\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \lnot(\textsf{entry\_present entryv}) }$
  $\specline{ \textsf{entryp+KERNBASE} \mapsto_{\mathsf{vpte,1}} \mathsf{entryp} \; \textsf{pfn\_set(entryv nextpaddr)}}$
  $\specline{\ghostmaptoken{\delta{}s}{\rtv}{\delta} \ast \begin{array}{l} \ulcorner\textsf{entry\_present (pte\_initialize(pfn\_set(entry\_val nextpaddr)),level)}\urcorner \wand \\  \;\;\;\;\;\;\; \forall_{i\in\textsf{0..511}} \; \textsf{table\_root (pte\_initialized (pfn\_set (entry\_val nextpaddr))) + i * 8} \mapsto_{\textsf{id}} \textsf{level-1}    \end{array}  }_{\rtv}$(*\label{line:initialize_post}*)
}
\end{lstlisting}
\vspace{-2em}
\caption{Allocating a physical page }
\label{pteinitializespecC}
\vspace{-1em}
\end{figure}

@@

  /*Similar steps & proofs for levels 3 and 2*/

@@

$\specline{\textsf{R}_{\textsf{walk}} \ast \textsf{R}_{\textsf{l1e}} }_{\rtv}$

@@

\begin{figure}\footnotesize
$\begin{array}{l}\left( \begin{array}{l} \ulcorner \textsf{l3}+\textsf{L3Offset(va)} = \textsf{l3\_entry} \land \textsf{table\_root(l4\_entry}_{\textsf{pfn}})  = \textsf{l3} \\ \land \textsf{l2}+\textsf{L2Offset(va)} = \textsf{l2\_entry} \land \textsf{table\_root(l3\_entry}_{\textsf{pfn}})  = \textsf{l2} \\ \land \textsf{l1}+\textsf{L1Offset(va)} = \textsf{l1\_entry} \land \textsf{table\_root(l2\_entry}_{\textsf{pfn}})  = \textsf{l1} \urcorner  \ast \\  \textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\setminus\{\textsf{l4\_entry},\textsf{l3\_entry},\textsf{l2\_entry},\textsf{l1\_entry}\},m)  \\ \ghostmaptoken{\delta{}s}{\rtv}{\delta} \ast  
      \ghostmaptoken{\textsf{id}}{(\mathsf{l4\_entry})}{\textsf{4}}\ast \ghostmaptoken{\textsf{id}}{(\mathsf{l3\_entry})}{\textsf{3}} \ast \ghostmaptoken{\textsf{id}}{(\mathsf{l2\_entry})}{\textsf{2}}  \ast \\  \textsf{l4\_entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{l4\_entry} \; \textsf{l4e\_val}  \ast \\  \textsf{l3\_entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{l3\_entry}\;\textsf{l3e\_val}  \ast \\  \textsf{l2\_entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{l2\_entry} \; \textsf{l2e\_val}  \end{array} \right)  =\textsf{R}_{\textsf{walk} } 
    \\ \ast \left(\begin{array}{l}  \textsf{l1\_entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{l1\_entry}\;\textsf{l1e\_val} \ast \\ \ulcorner  \textsf{qfrac} = 1 \leftrightarrow \; \lnot(\textsf{entry\_present l1e\_val})\urcorner \end{array}\right) =\textsf{R}_{\textsf{l1e}}\end{array} $
\todo[inline]{Colin: This is confusing: in the code \lstinline|l2_entry| is an actual entry word, but in these invariants they seem to be paddrs.}
\caption{\textsf{R}$_{\textsf{walk}}$: Resources obtained from invariant for walking page-table directory with \emph{non-existing entries} via calls to \textsf{l1e\_get\_next\_table} in Figure \ref{fig:calltopteinitializeC}}
\label{fig:rwalkC}
\vspace{-1em}
\end{figure}

@@

\todo[inline,color=cyan]{colin revision frontier}

\subsection{Mapping a New Page}
\label{sec:mapnewC}
One of the key tasks of a page fault handler in a general-purpose OS kernel is
to map new pages to an address space by writing into an existing page table via a call.\\
\centerline{\textsf{vaspace\_mappage(pte\_t *l4, void *va,uintptr\_t fpaddr)}}\\
in Figure \ref{fig:mapping_codeC}.
To do so, with a given allocated fresh page (\textsf{fpaddr}), then calculate the appropriate
known-valid page table walks (via \textsf{walkpgdir} Line \ref{line:call_walkpgdirC} in Figure \ref{fig:mapping_codeC}) and update
the appropriate L1 page table entry (Line 35 in Figure \ref{fig:mapping_codeC});
unmapping is the reverse of the logic we discuss here.
\looseness=-1
%\lstset{
%  columns=fullflexible,
%  numbers=left,
%  basicstyle=\ttfamily,
%  keywordstyle=\color{blue}\bfseries,
%  morekeywords={mov,add,call},
%  emph={rsp,rdx,rax,rbx,rbp,rsi,rdi,rcx,r8,r9,r10,r11,r12,r13,r14,r15},
%  emphstyle=\color{green},
%  emph={[2]cr3},
%  emphstyle={[2]\color{violet}},
%  morecomment=[l]{;;},
%  mathescape
%}
\begin{figure}\footnotesize
  \begin{lstlisting}[language=C,mathescape,escapeinside={(*}{*)}]
/* @param plm4: the virtual address of the root of the page table tree
   @param va: virtual address to be translated and mapped
   @param fpaddr: physical address of a zeroed page to map if va is not already mapped */
void vaspace_mappage(pte_t *l4, void *va,uintptr_t fpaddr ) {
  $\specline{\textsf{P} \ast \textsf{l4} \mapsto_{\textsf{id}} \textsf{\_} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \ghostmaptoken{\delta{}s}{\rtv}{\delta} \ast \ulcorner\theta \; !!\; \vaddr = \texttt{None}\urcorner}_{\rtv}$      
  pte_t *pteaddr = walkpgdir(l4, va); (*\label{line:call_walkpgdirC}*)
  $\specline{ \textsf{R}_{\textsf{walk}} \ast \textsf{R}_{\textsf{l1e}} }$
  if (!pteaddr->present){ (*\label{line:mappage_pte_present_startC}*)
    /*The entry is not present(*\label{line:mappage_pte_present_endC}*)*/
    pteaddr->pfn = PTE_ADDR_TO_PFN(fpaddr); /*Store updated entry back to L1 entry (*\label{line:l1entry_storeC}*)*/
    $\specline{ \textsf{l1\_entry+KERNBASE} \mapsto_{\textsf{vpte,1}} \textsf{l1\_entry} \; (\textsf{set\_pfn}(\textsf{l1e\_val}, \textsf{PTE\_ADDR\_TO\_PFN}(\textsf{fpaddr}))) \ast  \ulcorner \lnot(\textsf{entry\_present l1e\_val}) \urcorner \ast \textsf{P}}$
    /*Set the present bit in entry*/
    pteaddr->present = 1;
  }
  return;
  $\specline{ \begin{array}{l}\textsf{l1\_entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{l1\_entry} \; \textsf{l1e\_val} \ast \ulcorner (\textsf{entry\_present l1e\_val}) \urcorner \ast \textsf{P} \\
 \lor \; \;  \textsf{l1\_entry+KERNBASE} \mapsto_{\textsf{vpte,1}} \textsf{l1\_entry} \; (\textsf{set\_pfn}(\textsf{l1e\_val}, \textsf{PTE\_ADDR\_TO\_PFN}(\textsf{fpaddr}))) \ast  \ulcorner \lnot(\textsf{entry\_present l1e\_val}) \urcorner \ast P \end{array} }$
}
\end{lstlisting}
\vspace{-1em}
  \caption{Specification of updating L1 entry to reference a new page (\textsf{fpaddr}).}
\label{fig:mapping_codeC}
\end{figure}

In Figure \ref{fig:mapping_codeC}, we see an address ($\vaddr$) currently not
mapped to a page ($\theta \; !!\; \vaddr = \texttt{None}$). Mapping a fresh
physical page to back the desired virtual page first requires ensuring
the existence of a memory location for an appropriate L1 table entry.
The code uses a helper function \lstinline{walkpgdir} (discussed again in Section \ref{wlkpgdirC}).
\textsf{walkpgdir}'s postcondition contains virtual \emph{PTE} points-to assertions ($\mapsto_{\textsf{vpte}}$)
both for ensuring partial page table walk reaching the
L1 entry (l1e) by asserting that higher levels of the page table exist (R$_{\textsf{walk}}$ in Figure \ref{fig:rwalkC}), 
and for allowing access to the memory of the L1 entry via virtual address (R$_{\textsf{l1e}}$ in Figure \ref{fig:rwalkC}).

We already discussed for the upper level page-tables how the entry-present checks are handled.
However, for L1 entries this check is left to the caller of the
traversal function \textsf{walkpgdir}. In other words, unlike what we see in R$_{\textsf{walk}}$ for the upper levels, here all entry-present
checks have already been performed; the specification in R$_{\textsf{l1e}}$ ensures that page table entry for L1 needs to be checked at the caller site. 
In doing so, as we see in Figure \ref{fig:mapping_codeC}, the page reference \textsf{fpaddr} is linked back to the virtual address \textsf{va} 
only if it does not already refer to a physical resource (lines \ref{line:mappage_pte_present_startC}--\ref{line:mappage_pte_present_endC} in Figure \ref{fig:mapping_codeC}).

The crucial step in addition to traversing the page table in Figure \ref{walkpgdirC} is actually updating the L1 entry (Line \ref{line:updatepfnC} in Figure \ref{fig:mapping_codeC}),
via the virtual address (\textsf{pt\_entry+KERNBASE}) known to translate to the appropriate physical address, in our example the L1
table entry address ($\textsf{PTE\_ADDR\_TO\_PFN(fpaddr)}$).

Unlike the only prior work verifying analogous code for mapping a new page~\cite{kolanski08vstte,kolanski09tphols}, our proof above
does \emph{not} need to reason directly over the operational semantics,
making this the first verification we know of for mapping a virtual memory page that
stays entirely at the program logic level.
\looseness=-1
% By incorporating verification of the
% \lstinline|ensure_L1| function (see Section \ref{sec:traversing}), our verification also directly handles several subtle aspects which
% were axiomatized in prior work.
\ifPLDI
\else
\subsection{Unmapping a Page}
\todo[inline]{update (esp. line refs) for new mapping code}
The reverse operation, unmapping a designated page that is currently mapped,
would essentially be the reverse of
the reasoning around line 22 above: given the virtual points-to assertions for all 512
machine words of memory that the L1 entry would map,
and information about the physical location, 
full permission on the L1 entry could be obtained, allowing the construction of a
full virtual PTE pointer for it, setting to 0, and reclaiming the now unmapped physical memory.
\fi

>>>>

\begin{figure}\footnotesize
  \begin{lstlisting}[language=C,mathescape,escapeinside={(*}{*)}]
/* @param entry: virtual address of a non-present entry to initialize with a new physical page */
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \lnot(\textsf{entry\_present entryv}) \ast \textsf{entryp+KERNBASE} \mapsto_{\textsf{vpte,1}} \textsf{entryp entryv} \ast \ghostmaptoken{\textsf{id}}{(\mathsf{entryp})}{\textsf{level}} }$
void pte_initialize(pte_t *entry) {
  pte_t *local = kalloc(); // Allocate a full zeroed page for 512 8-byte entries(*\label{line:call_to_kallocC}*)
  entry->pfn = PTE_ADDR_TO_PFN((uintptr_t) local);(*\label{line:kalloc_install}*)
  $\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \lnot(\textsf{entry\_present entryv}) \ast \textsf{entryp+KERNBASE} \mapsto_{\mathsf{vpte,1}} \mathsf{entryp} \; \textsf{pfn\_set(entryv nextpaddr)}}$
  $\specline{\ghostmaptoken{\delta{}s}{\rtv}{\delta} \ast \begin{array}{l} \ulcorner\textsf{entry\_present (pte\_initialize(pfn\_set(entry\_val nextpaddr)),level)}\urcorner \wand \\  \;\;\;\;\;\;\; \forall_{i\in\textsf{0..511}} \; \textsf{table\_root (pte\_initialized (pfn\_set (entry\_val nextpaddr))) + i * 8} \mapsto_{\textsf{id}} \textsf{level-1}    \end{array}  }_{\rtv}$(*\label{line:initialize_post}*)
}
\end{lstlisting}
\vspace{-2em}
\caption{Allocating a physical page }
\label{pteinitializespecC}
\vspace{-1em}
\end{figure}

@@

@@

} $\specline{\textsf{R}_{\textsf{walk}} \ast \textsf{R}_{\textsf{l1e}} }_{\rtv}$

@@

\begin{figure}\footnotesize
$\begin{array}{l}\left( \begin{array}{l} \ulcorner \textsf{l3}+\textsf{L3Offset(va)} = \textsf{l3\_entry} \land \textsf{table\_root(l4\_entry}_{\textsf{pfn}})  = \textsf{l3} \\ \land \textsf{l2}+\textsf{L2Offset(va)} = \textsf{l2\_entry} \land \textsf{table\_root(l3\_entry}_{\textsf{pfn}})  = \textsf{l2} \\ \land \textsf{l1}+\textsf{L1Offset(va)} = \textsf{l1\_entry} \land \textsf{table\_root(l2\_entry}_{\textsf{pfn}})  = \textsf{l1} \urcorner  \ast \\  \textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\setminus\{\textsf{l4\_entry},\textsf{l3\_entry},\textsf{l2\_entry},\textsf{l1\_entry}\},m)  \\ \ghostmaptoken{\delta{}s}{\rtv}{\delta} \ast  
      \ghostmaptoken{\textsf{id}}{(\mathsf{l4\_entry})}{\textsf{4}}\ast \ghostmaptoken{\textsf{id}}{(\mathsf{l3\_entry})}{\textsf{3}} \ast \ghostmaptoken{\textsf{id}}{(\mathsf{l2\_entry})}{\textsf{2}}  \ast \\  \textsf{l4\_entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{l4\_entry} \; \textsf{l4e\_val}  \ast \\  \textsf{l3\_entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{l3\_entry}\;\textsf{l3e\_val}  \ast \\  \textsf{l2\_entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{l2\_entry} \; \textsf{l2e\_val}  \end{array} \right)  =\textsf{R}_{\textsf{walk} } 
    \\ \ast \left(\begin{array}{l}  \textsf{l1\_entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{l1\_entry}\;\textsf{l1e\_val} \ast \\ \ulcorner  \textsf{qfrac} = 1 \leftrightarrow \; \lnot(\textsf{entry\_present l1e\_val})\urcorner \end{array}\right) =\textsf{R}_{\textsf{l1e}}\end{array} $
\todo[inline]{Colin: This is confusing: in the code \lstinline|l2_entry| is an actual entry word, but in these invariants they seem to be paddrs.
Also, these should be 4 different fractions, not all qfrac
}
\caption{\textsf{R}$_{\textsf{walk}}$: Resources obtained from invariant for walking page-table directory with \emph{non-existing entries} via calls to \textsf{l1e\_get\_next\_table} in Figure \ref{fig:calltopteinitializeC}}
\label{fig:rwalkC}
\vspace{-1em}
\end{figure}

@@

\subsection{Mapping a New Page}
\label{sec:mapnewC}
Finally we come to the top-level routine for mapping a new page of memory into an address space
by updating page tables --- the \lstinline|vaspace_mappage| function in Figure \ref{fig:mapping_codeC}.
Again, our verification was carried out at the assembly level, but presented as on the original C for readability,
and the proof outline omits all relevant facts in favor of the most critial assertions involved in the key parts of the proof.
\lstinline|vaspace_mappage| is typically called by a page fault handler, to map a previously-reserved but lazily-allocated page.
It is passed the virtual address of the L4 table base (\lstinline|l4|), the virtual address to map (\lstinline|va|),
and the \emph{physical} address of an empty memory page which should be used as backing memory for \lstinline|va| and its surrounding page.
It begins by calling \lstinline|walkpgdir| (Line \ref{line:call_walkpgdirC}) to return the virtual address of 
the L1 entry which corresponds to \lstinline|va| (allocating intermediate tables as needed).
It then checks if the entry is already initialized. If not, \lstinline|fpaddr| is installed
into the L1 entry, which is then marked valid (setting the present bit), and the page is mapped.
% To do so, with a given allocated fresh page (\textsf{fpaddr}), then calculate the appropriate
% known-valid page table walks (via \textsf{walkpgdir} Line \ref{line:call_walkpgdirC} in Figure \ref{fig:mapping_codeC}) and update
% the appropriate L1 page table entry (Line 35 in Figure \ref{fig:mapping_codeC});
Unmapping is the reverse of the logic we discuss here.
\looseness=-1
%\lstset{
%  columns=fullflexible,
%  numbers=left,
%  basicstyle=\ttfamily,
%  keywordstyle=\color{blue}\bfseries,
%  morekeywords={mov,add,call},
%  emph={rsp,rdx,rax,rbx,rbp,rsi,rdi,rcx,r8,r9,r10,r11,r12,r13,r14,r15},
%  emphstyle=\color{green},
%  emph={[2]cr3},
%  emphstyle={[2]\color{violet}},
%  morecomment=[l]{;;},
%  mathescape
/* @param l4: the virtual address of the root of the page table tree
   @param va: virtual address to be translated and mapped
   @param fpaddr: physical address of a zeroed page to map if va is not already mapped */
void vaspace_mappage(pte_t *l4, void *va,uintptr_t fpaddr ) {
  $\specline{\textsf{P} \ast \textsf{l4} \mapsto_{\textsf{id}} \textsf{\_} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \ghostmaptoken{\delta{}s}{\rtv}{\delta} \ast \ulcorner\theta \; !!\; \vaddr = \texttt{None}\urcorner}_{\rtv}$ (*\label{line:mappage_pre}*)
  pte_t *pteaddr = walkpgdir(l4, va); (*\label{line:call_walkpgdirC}*)
  $\specline{ \textsf{R}_{\textsf{walk}} \ast \textsf{R}_{\textsf{l1e}} }$
  if (!pteaddr->present){ (*\label{line:mappage_pte_present_startC}*)
    /*The entry is not present(*\label{line:mappage_pte_present_endC}*)*/
    pteaddr->pfn = PTE_ADDR_TO_PFN(fpaddr); /*Store updated entry back to L1 entry (*\label{line:l1entry_storeC}*)*/
    $\specline{ \textsf{l1\_entry+KERNBASE} \mapsto_{\textsf{vpte,1}} \textsf{l1\_entry} \; (\textsf{set\_pfn}(\textsf{l1e\_val}, \textsf{PTE\_ADDR\_TO\_PFN}(\textsf{fpaddr}))) \ast  \ulcorner \lnot(\textsf{entry\_present l1e\_val}) \urcorner \ast \textsf{P}}$
    /*Set the present bit in entry*/
    pteaddr->present = 1;(*\label{line:l1entry_setpresent}*)
  $\specline{ \begin{array}{l}\textsf{l1\_entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{l1\_entry} \; \textsf{l1e\_val} \ast \ulcorner (\textsf{entry\_present l1e\_val}) \urcorner \ast \textsf{P} \\
 \lor \; \;  \textsf{l1\_entry+KERNBASE} \mapsto_{\textsf{vpte,1}} \textsf{l1\_entry} \; (\textsf{set\_pfn}(\textsf{l1e\_val}, \textsf{PTE\_ADDR\_TO\_PFN}(\textsf{fpaddr}))) \ast  \ulcorner \lnot(\textsf{entry\_present l1e\_val}) \urcorner \ast P \end{array} }$
  \caption{Specification of updating L1 entry to reference a new page (\textsf{fpaddr}).}
\label{fig:mapping_codeC}

For brevity this example is specialized to the case where $\vaddr$ is known to not be mapped:
the precondition on Line \ref{line:mappage_pre} includes $\theta \; !!\; \vaddr = \texttt{None}$.\footnote{
  The generalization to handling either case and returning an error is straightforward.
}
\todo[inline,color=cyan]{
  Ismail, shouldn't the precondition include the 512 tokens for the L4 table, and shouldn't the l4 entry be excluded from $\Xi$?
}
The precondition directly implies the precondition of \lstinline|walkpgdir|, which as just discussed
returns the virtual address of an allocated L1 entry, and whose postcondition contains almost all of the information
needed to construct a virtual points-to for \lstinline|va| --- except information about the L1 entry
being present and pointing to a data page.
We already discussed for the upper level page-tables how the entry-present checks are handled, and
Line \ref{line:mappage_pte_present_startC} is similar: $\mathsf{R}_{\textsf{l1e}}$ includes the self-conditional
fractional permission for the L1 entry, so as it is not present, by Line \ref{line:l1entry_storeC}
it is known that full permission is held to update that entry.
Lines \ref{line:l1entry_storeC} and \ref{line:l1entry_setpresent} are verified using the pte-points-to
memory store rule.\footnote{In the assembly, \lstinline|fpaddr| is already known to be word-aligned, so it is simply bitwise-and-ed with 1
(setting the present flag) and a single store occurs.}
\looseness=-1

\todo[inline]{Colin will fix up this paragraph: it's right but also kind of weird, will handle during related work pass}
Unlike the only prior work verifying analogous code for mapping a new page~\cite{kolanski08vstte,kolanski09tphols}, our proof above
does \emph{not} need to reason directly over the operational semantics,
making this the first verification we know of for mapping a virtual memory page that
stays entirely at the program logic level.
\looseness=-1
% By incorporating verification of the
% \lstinline|ensure_L1| function (see Section \ref{sec:traversing}), our verification also directly handles several subtle aspects which
% were axiomatized in prior work.
\ifPLDI
\else
\subsection{Unmapping a Page}
\todo[inline]{update (esp. line refs) for new mapping code}
The reverse operation, unmapping a designated page that is currently mapped,
would essentially be the reverse of
the reasoning around line 22 above: given the virtual points-to assertions for all 512
machine words of memory that the L1 entry would map,
and information about the physical location, 
full permission on the L1 entry could be obtained, allowing the construction of a
full virtual PTE pointer for it, setting to 0, and reclaiming the now unmapped physical memory.
\fi
